Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-13 _manuscript/_tex/index.tex
illustration and cite veridical ds

The externalization of the data analysis process is a practice that has
potential to improve the trustworthiness of analyses in general and the
trustworthiness of subsequent products, such as machine learning models,
that may be built on such analyses. While publication of analysis code
and data is now a common a requirement for the sake of reproducibility
\citep{peng2011reproducible}, the publication code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data processing. Thus, a reader looking at
published code can often be left with many questions about why certain
choices were made. Developing an approach to reveal some of this process
without requiring a reader to essentially reconstruct the analysis
process from the raw data would provide an improved basis for trusting
an analysis result \citep{peng2021reproducible}.

@@

\includegraphics[width=6.73in,height=\textheight]{figures/metric-calc.png}

@@

\caption{\label{fig-metric-calc}{[}this is the cap and the plot needs
polish{]}}

>>>>

The externalization of the data analysis process is a practice that has
potential to improve the trustworthiness of analyses in general
\citep{yu2024veridical62} and the trustworthiness of subsequent
products, such as machine learning models, that may be built on such
analyses. While publication of analysis code and data is now a common a
requirement for the sake of reproducibility
\citep{peng2011reproducible}, the publication code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data processing. Thus, a reader looking at
published code can often be left with many questions about why certain
choices were made. Developing an approach to reveal some of this process
without requiring a reader to essentially reconstruct the analysis
process from the raw data would provide an improved basis for trusting
an analysis result \citep{peng2021reproducible}.

@@

\includegraphics[width=1\textwidth,height=\textheight]{figures/metric-calc.png}

@@

\caption{\label{fig-metric-calc}Overview of the procedure to quantify
the effectiveness of analysis validation checks. The procedure involves
simulating replicates of the data, applying the anlysis, and running the
analysis validation checks on each replicate to determine whether the
outcome meets the expectation (the unexpected column) and whether each
check passes (check 1, check 2, \ldots). The simulation results are then
passed to two branches: the accuracy branch calculates the precision and
recall of the checks from a logic regression prediction, while the
redundancy branch to calculate the mutual information and independence
of the checks. These three metrics are combined to produce a single
metric that quantifies the effectiveness of the checks in diagnosing
unexpected outcomes in the analysis.}

### 2024-12-13 index.qmd
illustration and cite veridical ds

The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing. Thus, a reader looking at published code can often be left with many questions about why certain choices were made. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

@@

```{r}
#| label: fig-metric-calc
#| fig-cap: "[this is the cap and the plot needs polish]"
knitr::include_graphics(here::here("figures", "metric-calc.png"))
```

>>>>

The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical62] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses. While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis. Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing. Thus, a reader looking at published code can often be left with many questions about why certain choices were made. Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].

@@

```{r}
#| label: fig-metric-calc
#| out.width: 100%
#| fig-cap: "Overview of the procedure to quantify the effectiveness of analysis validation checks. The procedure involves simulating replicates of the data, applying the anlysis, and running the analysis validation checks on each replicate to determine whether the outcome meets the expectation (the unexpected column) and whether each check passes (check 1, check 2, ...). The simulation results are then passed to two branches: the accuracy branch calculates the precision and recall of the checks from a logic regression prediction, while the redundancy branch to calculate the mutual information and independence of the checks. These three metrics are combined to produce a single metric that quantifies the effectiveness of the checks in diagnosing unexpected outcomes in the analysis."
knitr::include_graphics(here::here("figures", "metric-calc.png"))
```
